### 2020-04-10 Paper_shortened_pdf.Rmd
Final changes for resubmission

---
title             : "Quantitative involvement moderates the within-family interplay between mothers' and fathers' qualitative parenting and children's emotional adjustment"
shorttitle        : "INVOLVEMENT AND EMOTIONAL ADJUSTMENT"

@@

abstract: |
  Ample research has shown that children's emotional adjustment and parenting
  are related. Several important questions remain unaddressed. First, both
  quantitative involvement (time spent on childcare), and qualitative
  involvement (specific parenting behaviors) have been implicated in emotional
  adjustment development. Yet these involvement dimensions are rarely
  investigated jointly or in interaction. Secondly, although fathers have been
  theorized to contribute uniquely to emotional adjustment development, they
  are not routinely included in research. Consequently, the potential
  implications of joint- (total) and relative (who does what) quantitative
  involvement for the interplay between qualitative parenting and emotional
  adjustment have not been adequately researched. This longitudinal study of
  Australian families (N = 1884, from age 6-12) addressed these questions,
  using dynamic structural equation modeling to examine the within-family
  interplay between qualitative parenting and emotion regulation, while
  controlling for stable between-family differences in these constructs. Joint
  and relative quantitative involvement, and child sex, were included as
  between-family moderators of within-family processes. We primarily found
  between-family correlations between qualitative involvement and emotional
  adjustment. Evidence for within-family effects was limited: On average,
  maternal warmth positively predicted emotional adjustment. Several child
  effects indicated that emotional adjustment evokes changes in parenting.
  Within-family effects varied substantially across families however,
  especially for fathers. When joint quantitative involvement was higher, both
  parents engaged in more desirable qualitative parenting. When mothers were
  relatively more involved than fathers, parental influence was more
  differentiated, fathers displayed less desirable qualitative parenting, and
  children were less well-adjusted. This suggests that all family members
  benefit from a more equal childcare distribution.

@@

\vspace{\baselineskip}
Children's emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for their child), and in terms of qualitative parenting behaviors (how parents behave towards their child). Nevertheless, the existing literature has two major shortcomings. Firstly, although it is widely acknowledged that quantitative and qualitative involvement are both relevant to child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004], these dimensions are rarely jointly investigated, let alone in interaction [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. Therefore, we currently do not know how the interplay between quantitative and qualitative involvement relates to emotional development. For example, children's emotional adjustment might be differently impacted by anger coming from a highly involved parent, than from a generally absent parent. Conversely, children's poor emotional adjustment might elicit warm responses from hands-on parents, who are aware of the reasons for their children's emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be disciplined with anger or inductive reasoning. The present study overcomes this limitation by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.

@@

Early developmental research focused on parental socialization; the notion that parents shape child development through proximal mechanisms, such as qualitative parenting behaviors [@maccobySocializationContextFamily1983; see also @vanlissaRoleFathersMothers2018a]. Qualitative parenting behaviors are related to the dimensions underlying parenting styles: Responsiveness and demandingness [@maccobySocializationContextFamily1983]. Substantial evidence indicates that responsiveness, which we operationalized in terms of warm and supportive parenting, is positively associated with children's emotional adjustment [@morrisRoleFamilyContext2007]. Therefore we hypothesized that parental warmth would be positively associated with children's emotional adjustment, $H_{1a}$.<!--@khalequePerceivedParentalAcceptanceRejection2002; -->

@@

A view of the family as a dynamic system, whose members reciprocally influence one another, has become commonplace [@kerrPerceivedParentingStyle2012]. Nevertheless, bi-directional effects are not routinely investigated in the childhood literature [@flouriFathersBehaviorsChildren2010]. Consequently, most of the research cited in support of hypotheses $H_{1a-d}$ is cross-sectional, and could be taken as support for child effects as well as parent effects. We therefore hypothesize that child effects are similar in direction, and that emotional adjustment positively predicts warmth, consistency, and inductive reasoning, and negatively predicts anger, $H_{2a-d}$. This is supported by studies that investigated child effects specifically, and found that children's maladjustment elicits less parental warmth, and more negative parenting [@gadeyneLongitudinalRelationsParenting2004]. Bi-directional positive links between parental consistency and emotional adjustment have also been reported [@lenguaBidirectionalAssociationsTemperament2005]. 
The adolescent literature provides additional evidence that emotional adjustment positively predicts parental support [@otterpohlCrossLaggedRelationsParenting2015; @vandergiessenPerceivedAutonomySupport2014], and negatively predicts negative parenting [@wijsbroekDirectionEffectsPerceived2011]. Thus, in a way, parents respond positively to emotional adjustment, and negatively to maladjustment.
<!-- Neppl 2009: Adolescents' harsh behavior towards parents predicts harsh parenting, and positive behavior predicts positive parenting -->

@@

Methods commonly used to examine direction of effects between parents and children have been criticized for yielding biased results if stable between-family differences are not controlled [@hamakerCritiqueCrosslaggedPanel2015]. Innovative methods provide better estimates of dynamic reciprocal processes within families, by controlling for stable between-family differences: The recently developed random-intercept cross-lagged panel model, which is an extension of the standard cross-lagged panel model [@hamakerCritiqueCrosslaggedPanel2015], and dynamic structural equation modeling (DSEM); a multilevel generalization of this model [@asparouhovDynamicStructuralEquation2018].

Several studies have re-examined the direction of effects between parents and children using these methods, but (to our knowledge) only in adolescent samples. One such study found that effects of parenting on adjustment that emerged using classical methods, disappeared when stable between-family differences in parenting and adjustment were included in the model [@keijsersParentalMonitoringAdolescent2016]. Another study similarly found that most associations between maternal and paternal qualitative parenting and adolescents' emotional adjustment, were explained by between-family level correlations. The remaining predictive effects consisted primarily of child effects, not parenting effects [@vanlissaRoleFathersMothers2018a]. The authors theorized that this finding was reflective of the relative equality of parent-child relationships in adolescence. If this is correct, we would expect to find more evidence of parenting effects in the present, younger, sample. These two studies also illustrate that applying these new methods changes the pattern of results in unexpected ways. Consequently, it is impossible to make specific predictions about the pattern of between-family correlations and within-family predictive effects. For now, we will assume that $H_{1a-d}$ and $H_{2a-d}$ apply at both the between- and within-family level.

@@

In line with this theory, empirical research has found that mothers' responses to emotional expression are more warm and supportive than fathers' responses, which are more demanding and punitive [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. In adolescence, maternal warmth was found to be a more salient socialization mechanism of emotional adjustment than maternal demandingness, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2018a]. Similar differentiated associations were found for younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Finally, there is initial evidence that emotional adjustment is more negatively impacted by mothers' negative parenting, compared to fathers [@changHarshParentingRelation2003]; perhaps because negative parenting is somewhat incongruous with the warm "attachment" relationship typically expected of mothers. Based on this research, we therefore hypothesize that emotional adjustment would be more strongly associated with maternal warmth and anger, and with paternal demandingness, operationalized as consistency and inductive reasoning, $H_{3}$.

@@

Early fatherhood research primarily operationalized involvement in *quantitative* terms: Presence versus absence in divorced families, and the sharing of childcare duties in dual-earner families [@lambBiosocialPerspectivePaternal1987]. Quantitative involvement has been researched extensively in terms of the absolute amount of childcare provided [see @pleckPaternalInvolvementRevised2004]. However, some have argued that the *division* of childcare tasks is more relevant for children's adjustment [@krollEarlyFatherInvolvement2016; @sabattiniRelationMothersFathers2004; @craigHowMothersFathers2011; @feinbergInternalStructureEcological2003]. Indeed, the equity of the childcare distribution is associated with socio-emotional development [@deutschPaternalParticipationChild2001]. Moreover, an unequal distribution often implies that women perform a disproportionate share of the work [@craigDoesFatherCare2006]. This strains the marital relationship [@shockleyItNotWhat2018], and erodes parenting quality, particularly for parents of difficult children [@sheltonInterparentalConflictNegative2008].

There are also methodological advantages to focusing on the distribution of childcare. Operationalizing quantitative involvement in terms of both parents' absolute involvement incurs a potential confound: Effects that appear to be due to fathers' greater involvement might be due to children receiving *more* total care [see @deutschPaternalParticipationChild2001]. This confound can be mitigated by examining the effect of the distribution of childcare, while controlling for total care. Some prior research has attempted to achieve this by assessing the childcare distribution directly, by asking which parent is most responsible for certain tasks [@krollEarlyFatherInvolvement2016; @keizerInfluenceFathersMothers2019].

Alternatively, when quantitative involvement has been measured in terms of absolute involvement, one can estimate a latent mean and difference to statistically disentangle joint involvement from relative involvement [@mcardleLatentVariableModeling2009]. Joint involvement reflects the total (average) amount of care provided to children, and relative involvement reflects the extent to which the mother is more involved than the father (i.e., the childcare distribution).

@@

The amount of time parents spend with children is associated with the quality of parenting behaviors [@sabattiniRelationMothersFathers2004]. Indeed, more involved parents tend to experience greater parenting self-efficacy, possibly reflecting the effects of experience [@gialloMothersFathersInvolvement2013a]. Furthermore, when both parents are more involved, they monitor each other's parenting behaviors to ensure that children are raised well [@mclanahanConsequencesFatherAbsence2014]. Consequently, we hypothesize that greater joint involvement will positively predict both parents' desirable parenting behaviors (warmth, consistency, and inductive reasoning), and negatively predict parental anger, $H_{4a-d}$.

@@

Bioecological theories of development indicate that children require a certain amount of care, and who provides this care is, to some extent, flexible [see @bronfenbrennerEcologyHumanDevelopment1979]. We therefore hypothesize that greater joint care should positively predict children's emotional adjustment, $H_{6a}$. Further, it has been argued that children feel less supported when they perceive that one of their parents is relatively uninvolved, than when both parents are equally invested in their well-being [@deutschPaternalParticipationChild2001]. We therefore predict that greater relative involvement will negatively predict emotional adjustment, $H_{6b}$.

@@

Child sex differences are considered to be an important covariate in the literature on parenting and emotional adjustment. Parents raise boys and girls differently, expressing more demandingness with girls [@pomerantzRoleMaternalControl1998; @vanlissaRoleFathersMothers2018a], and more anger with boys [@strausCorporalPunishmentAmerican1999]. Furthermore, the same-sex parent is thought to play a stronger role in socialization [@busseySocialCognitiveTheory1999]. Finally, mounting evidence suggests that parental warmth is greatest in mother-daughter relationships, and lowest in father-son relationships [see @vanlissaRoleFathersMothers2018a]. We therefore hypothesize that parents will display more warmth, consistency, and inductive reasoning, and less anger, towards girls than boys, $H_{7a-d}$. We also predict associations between qualitative parenting and emotional adjustment to be stronger between children and their same-sex parent, $H_{8a-d}$. Finally, although sex differences in emotional adjustment have been found in older children [@vanlissaRoleFathersMothers2018a], there is little evidence for such differences in young children [@else-questGenderDifferencesTemperament2006a].

@@

The present study examined whether specific qualitative parenting behaviors are associated with children's emotional adjustment, and whether these associations are moderated by quantitative involvement. We aimed to make several contributions. First, by including both mothers and fathers, we can tease apart their common and unique roles in children's emotional adjustment development. Second, by incorporating both qualitative and quantitative measures of involvement, we can examine interactions between qualitative and quantitative involvement, which have been largely overlooked in prior work. Third, by disentangling stable between-family differences from dynamic within-family predictive processes, we can more accurately assess the direction of effects between qualitative involvement and emotional adjustment. Finally, we examined child sex as a moderator of between-family differences in all study variables and within-family predictive effects.

@@

We calculated mean scores for all study variables with multiple indicators. Scale descriptive statistics, number of items, attrition (which ranged from 7-14% across respondents), and reliability are reported in Supplementary Table 1, and scale correlations are provided in Supplementary Table 2. The difference in fit between structurally invariant models and metrically invariant models was smaller than .01 on all conventional fit indices: RMSEA, CFI, and TLI (see Supplementary Table 3). We thus deemed it defensible to assume metric measurement invariance (i.e., to assume that the factor loadings of each construct are the same across time points and respondents).

@@

To address our research questions, we used multilevel, autoregressive structural equation modeling [@asparouhovDynamicStructuralEquation2018]. In this approach, the variance in qualitative parenting behaviors and emotional adjustment is partialized into stable between-individual differences (the mean level for that individual across all measurement occasions), and within-individual fluctuations (deviations from this person-specific mean). This has several advantages: First, controlling for stable between-family differences allows us to estimate the direction of within-family effects more accurately [@hamakerCritiqueCrosslaggedPanel2015]. Any stable factors which might bias cross-lagged effects - such as the influence of SES or neighborhood effects - are accounted for at the between level, and reflected in the correlations between the random intercepts. Secondly, this approach enables us to directly test our hypothesis that between-family differences in quantitative involvement moderate the interplay between qualitative involvement and emotional adjustment: Between-family differences in the interplay between qualitative parenting behaviors and emotional adjustment are estimated as random effects, and we can include between-family level moderators as predictors of these random effects. In this way, we included joint and relative quantitative involvement, and child sex as moderators. Figure \@ref(fig:figure1) shows a graphical representation of the resulting model. We conducted separate analyses for the four qualitative parenting dimensions, because a model including all four variables at once was too complex to estimate.

@@

We hypothesized that warmth, consistency, and inductive reasoning would correlate positively, and anger negatively, with emotional adjustment. At the between-family level, we examined correlations between families' average levels of qualitative parenting behaviors and emotional adjustment (see Table \@ref(tab:tabcors)). These correlations indicate whether between-family differences in the average levels of these parenting behaviors are associated with between-family differences in children's emotional adjustment. Counter to hypotheses $H_{1,2a}$, we found no significant correlations between parental warmth and children's emotional adjustment. In support of hypotheses $H_{1,2b}$, we found significant positive correlations between maternal and paternal consistency, and emotional adjustment. Contrary to hypotheses $H_{1,2c}$, we found that maternal and paternal inductive reasoning were negatively correlated with emotional adjustment. Finally, as hypothesized in $H_{1,2d}$, we found significant negative correlations between maternal and paternal anger, and emotional adjustment. This means that, in families where parents reported more consistency, children displayed better emotional adjustment - and in families where parents reported more inductive reasoning or anger, children showed poorer emotional adjustment.

@@

We also examined within-family correlations, which are based on within-family fluctuations in the study variables (Table \@ref(tab:tabcors)). These correlations indicate whether fluctuations in parents' parenting behaviors are correlated with fluctuations in their children's emotional adjustment. In support of hypothesis $H_{1a}$, we found significant positive within-family level associations between fathers' and mothers' warmth, and emotional adjustment. This means that if, in any particular wave, parents report more warmth than usual, their children displayed better emotional adjustment than usual. Furthermore, in partial support of hypotheses $H_{1b,c}$, we found significant positive correlations of emotional adjustment with  mothers', but not fathers', consistency, as well as with fathers', but not mothers', inductive reasoning. Finally, contrary to hypothesis $H_{1d}$, we found no significant within-family associations between children's emotional adjustment and maternal or paternal anger. Counter to hypothesis $H_{3}$, none of the differences between mothers and fathers in the strength of either between- or within-family level correlations were significant (see Supplementary Table S4). This includes the within-family correlation between warmth and emotional adjustment, whose effect size was nearly twice as large for fathers as compared to mothers.

## Predictive effects between within-family changes in parenting and emotional adjustment

@@

With regard to child effects, in line with hypothesis $H_{2a}$, the average cross-lagged effect of emotional adjustment on maternal warmth was positive and significant (Table \@ref(tab:tabcrosspath)). Thus, on average, when children displayed better emotional adjustment than usual in one wave, their mothers displayed more warmth than usual in the next wave. Contrary to hypothesis $H_{2b}$, the average effect of emotional adjustment on maternal consistency was non-significant. As hypothesized in $H_{2c}$, the intercept for maternal inductive reasoning was positive. Counter to hypothesis $H_{2d}$, the intercept for maternal anger was positive, not negative. Thus, when children displayed better emotional adjustment than usual in one wave, their mothers displayed more anger than usual in the next wave. In support of hypothesis $H_{2a}$, the average child effect of adjustment on paternal warmth was positive and significant. None of the other intercepts ($H_{2b-d}$) for child effects on paternal parenting were significant. The variances of all child effects were significant but small, indicating that the effect of children's emotional adjustment on their parents' child-rearing behaviors differs somewhat between families.

@@

We conducted exploratory analyses of interactions between involvement quality and quantity, by investigating whether between-family differences in quantitative involvement had significant effects on the within-family cross-lagged effects between qualitative parenting behaviors and emotional adjustment (Table \@ref(tab:tabinvolvement)). The novelty of this research area precluded us from formulating specific hypotheses. With regard to parenting effects, the only significant moderation we found was for maternal warmth. Specifically, in families where both parents were more involved, the cross-lagged effect of maternal warmth on emotional adjustment was *lower* than in families with less-involved parents. Specifically, $+1SD$ in joint involvement predicted a $-.14SD$ decrease in the within-families cross-lagged effect of maternal warmth on emotional adjustment (see Table \@ref(tab:tabcrosspath)). This same within-families effect increased by $.11$ of a standard deviation of the effect when relative involvement was one standard deviation higher, i.e., when mothers were more involved than fathers. We found no moderation for paternal warmth, nor for both parents' consistency, inductive reasoning, and anger.

@@

## Effects of child sex

>>>>

---
title             : "Mothers' and fathers' quantitative and qualitative parenting in relation to children's emotional adjustment: A between- and within-family investigation"
shorttitle        : "INVOLVEMENT AND EMOTIONAL ADJUSTMENT"

@@

abstract: |
  This longitudinal study of Australian families (n = 1884, from age 6-12)
  examined how fathers' and mothers' quantitative involvement (time spent on
  childcare) and qualitative involvement (specific parenting behaviors) relate
  to children's emotional adjustment development. We used dynamic structural
  equation modeling to disentangle stable between-family differences from
  within-family fluctuations in qualitative parenting and emotional adjustment,
  allowing us to investigate the direction of effects between parents and
  children. Because fathers have been theorized to contribute uniquely to
  emotional adjustment development, we examined differences between mothers and
  fathers, and the interplay between parent- and child sex. We further examined
  whether between-family differences in quantitative involvement,
  operationalized as joint (total) and relative (one parent does more)
  involvement, predict qualitative parenting behaviors and emotional
  adjustment, and moderate the within-family interplay between them. Our
  results primarily revealed between-family correlations of qualitative
  involvement with emotional adjustment. Evidence for within-family effects was
  limited to a parenting effect of maternal warmth, and several child effects,
  which indicated that emotional adjustment evokes changes in parenting.
  Parenting effects varied substantially across families however, especially
  for fathers. Between-family differences in quantitative involvement explained
  some of these differences. When joint quantitative involvement was higher,
  both parents engaged in more desirable qualitative parenting. When mothers
  were relatively more involved than fathers, fathers displayed less desirable
  qualitative parenting, and children were less well-adjusted. There was some
  limited evidence that within-family predictive effects were moderated by
  quantitative involvement, particularly relative involvement. These findings
  suggest that all family members benefit from a more equal childcare
  distribution.

@@

\vspace{\baselineskip}
Children's emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for their child), and in terms of qualitative parenting behaviors (how parents behave towards their child). Nevertheless, the existing literature has two major shortcomings. Firstly, although it is widely acknowledged that quantitative and qualitative involvement are both relevant to child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004], these dimensions are rarely jointly investigated, let alone in interaction [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. Yet it is plausible that quantitative involvement might moderate predictive effects between qualitative parenting and emotional adjustment. For example, children's emotional adjustment might be differently impacted by anger coming from a highly involved parent, than from a generally absent parent. Conversely, children's poor emotional adjustment might elicit warm responses from hands-on parents, who are aware of the reasons for their children's emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be disciplined with anger or inductive reasoning. The present study overcomes this limitation by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.

@@

Early developmental research focused on parental socialization; the notion that parents shape child development through proximal mechanisms, such as qualitative parenting behaviors [@maccobySocializationContextFamily1983; see also @vanlissaRoleFathersMothers2019]. Qualitative parenting behaviors are related to the dimensions underlying parenting styles: Responsiveness and demandingness [@maccobySocializationContextFamily1983]. Substantial evidence indicates that responsiveness, which we operationalized in terms of warm and supportive parenting, is positively associated with children's emotional adjustment [@morrisRoleFamilyContext2007]. Therefore we hypothesized that parental warmth would be positively associated with children's emotional adjustment, $H_{1a}$.<!--@khalequePerceivedParentalAcceptanceRejection2002; -->

@@

The family is commonly viewed as a dynamic system, whose members reciprocally influence one another [@kerrPerceivedParentingStyle2012]. Nevertheless, bi-directional effects are not routinely investigated [@flouriFathersBehaviorsChildren2010]. Most of the research cited in support of hypotheses $H_{1a-d}$ is cross-sectional, and could also support child effects. We thus hypothesize that emotional adjustment positively predicts warmth, consistency, and inductive reasoning, and negatively predicts anger, $H_{2a-d}$. These predictions are in line with studies that investigated child effects specifically, and found that maladjustment elicits less warmth and more negative parenting [@gadeyneLongitudinalRelationsParenting2004]. There is also evidence for bi-directional positive links between parental consistency and emotional adjustment [@lenguaBidirectionalAssociationsTemperament2005]. The adolescent literature provides additional evidence that emotional adjustment positively predicts parental support [@otterpohlCrossLaggedRelationsParenting2015; @vandergiessenPerceivedAutonomySupport2014], and negatively predicts negative parenting [@wijsbroekDirectionEffectsPerceived2011]. Thus, in a way, parents respond positively to emotional adjustment, and negatively to maladjustment.
<!-- Neppl 2009: Adolescents' harsh behavior towards parents predicts harsh parenting, and positive behavior predicts positive parenting -->

@@

Methods commonly used to examine direction of effects between parents and children have been criticized for failing to disentangle between-family differences from within-family predictive effects [@hamakerCritiqueCrosslaggedPanel2015]. The recently developed random-intercept cross-lagged panel model, which is an extension of the standard cross-lagged panel model [@hamakerCritiqueCrosslaggedPanel2015], and dynamic structural equation modeling (DSEM); a multilevel generalization of this model [@asparouhovDynamicStructuralEquation2018], overcome this limitation by disentangling the two levels.

Several studies have re-examined the direction of effects between parents and children using these methods, but mostly in adolescent samples. One such study found that effects of parenting on adjustment that emerged using classical methods became non-significant when stable between-family differences were taken into account [@keijsersParentalMonitoringAdolescent2016]. Another study similarly found that associations between maternal and paternal qualitative parenting and adolescents' emotional adjustment, were mostly explained by between-family level correlations. The remaining predictive effects consisted primarily of child effects, not parenting effects [@vanlissaRoleFathersMothers2019]. The authors theorized that this finding was reflective of the relative equality of parent-child relationships in adolescence. If this is correct, we would expect to find more evidence of parenting effects in the present, younger, sample. These two studies also illustrate that applying these new methods changes the pattern of results in unexpected ways. Consequently, it is impossible to make specific predictions about the pattern of between-family correlations and within-family predictive effects. For now, we will assume that $H_{1a-d}$ and $H_{2a-d}$ apply at both the between- and within-family level.

@@

In line with this theory, empirical research has found that mothers' responses to emotional expression are more warm and supportive than fathers' responses, which are more demanding and punitive [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. In adolescence, maternal warmth was found to be a more salient socialization mechanism of emotional adjustment than maternal demandingness, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2019]. Similar differentiated associations were found for younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Finally, there is initial evidence that emotional adjustment is more negatively impacted by mothers' negative parenting, compared to fathers [@changHarshParentingRelation2003]; perhaps because negative parenting is somewhat incongruous with the warm "attachment" relationship typically expected of mothers. Based on this research, we therefore hypothesize that emotional adjustment would be more strongly associated with maternal warmth and anger, and with paternal demandingness, operationalized as consistency and inductive reasoning, $H_{3}$.

@@

Early fatherhood research primarily operationalized involvement in *quantitative* terms: Father absence in divorced families, and childcare sharing in dual-earner families [@lambBiosocialPerspectivePaternal1987]. Quantitative involvement has been researched extensively in terms of the absolute amount of childcare provided [see @pleckPaternalInvolvementRevised2004]. However, some have argued that the *division* of childcare tasks is more relevant for children's adjustment [e.g., @keizerInfluenceFathersMothers2019; @craigHowMothersFathers2011]. Indeed, the equity of the childcare distribution is associated with socio-emotional development [@deutschPaternalParticipationChild2001]. Moreover, an unequal distribution often implies that women perform a disproportionate share of the work [@craigDoesFatherCare2006]. This strains the marital relationship [@shockleyItNotWhat2018], and erodes parenting quality, particularly for parents of difficult children [@sheltonInterparentalConflictNegative2008].

There are also methodological advantages to focusing on the distribution of childcare. Operationalizing quantitative involvement in terms of both parents' absolute involvement incurs a potential confound: Effects that appear to be due to fathers' greater involvement might be due to children receiving *more* total care [see @deutschPaternalParticipationChild2001]. This confound can be mitigated by estimating a latent mean and difference [@mcardleLatentVariableModeling2009]. The latent mean reflects joint involvement; the total (average) amount of care provided to children, and tge latent difference reflects relative involvement; the extent to which the mother is more involved than the father (i.e., the childcare distribution).

<!-- examining the effect of the distribution of childcare, while controlling for total care. Some prior research has attempted to achieve this by assessing the childcare distribution directly, by asking which parent is most responsible for certain tasks [@krollEarlyFatherInvolvement2016; ].

Alternatively, when quantitative involvement has been measured in terms of absolute involvement, one can-->

@@

Time spent with children is associated with the quality of parenting behavior [@sabattiniRelationMothersFathers2004]. Indeed, more involved parents tend to experience greater parenting self-efficacy, possibly reflecting the effects of experience [@gialloMothersFathersInvolvement2013a]. Furthermore, when both parents are more involved, they monitor each other's parenting behaviors to ensure that children are raised well [@mclanahanConsequencesFatherAbsence2014]. Consequently, we hypothesize that greater joint involvement will positively predict both parents' desirable parenting behaviors (warmth, consistency, and inductive reasoning), and negatively predict parental anger, $H_{4a-d}$.

@@

Bioecological theories argue that children require a certain amount of care; who provides it is somewhat flexible [see @bronfenbrennerEcologyHumanDevelopment1979]. We therefore hypothesize that greater joint care should positively predict children's emotional adjustment, $H_{6a}$. Further, it has been argued that children feel less supported when they perceive that one of their parents is relatively uninvolved, than when both parents are equally invested in their well-being [@deutschPaternalParticipationChild2001]. We therefore predict that greater relative involvement will negatively predict emotional adjustment, $H_{6b}$.

@@

Child sex differences are considered to be an important covariate in the literature on parenting and emotional adjustment. Parents raise boys and girls differently, expressing more demandingness with girls [@pomerantzRoleMaternalControl1998; @vanlissaRoleFathersMothers2019], and more anger with boys [@strausCorporalPunishmentAmerican1999]. Furthermore, the same-sex parent is thought to play a stronger role in socialization [@busseySocialCognitiveTheory1999]. Finally, mounting evidence suggests that parental warmth is greatest in mother-daughter relationships, and lowest in father-son relationships [see @vanlissaRoleFathersMothers2019]. We therefore hypothesize that parents will display more warmth, consistency, and inductive reasoning, and less anger, towards girls than boys, $H_{7a-d}$. We also predict associations between qualitative parenting and emotional adjustment to be stronger between children and their same-sex parent, $H_{8a-d}$. Finally, although sex differences in emotional adjustment have been found in older children [@vanlissaRoleFathersMothers2019], there is little evidence for such differences in young children [@else-questGenderDifferencesTemperament2006a].

@@

The present study examined whether specific qualitative parenting behaviors are associated with children's emotional adjustment, and whether these associations are moderated by quantitative involvement. We aimed to make several contributions. First, by including both mothers and fathers, we can tease apart their common and unique roles in children's emotional adjustment development. Second, by incorporating both qualitative and quantitative measures of involvement, we can examine interactions between qualitative and quantitative involvement, which have been largely overlooked in prior work. Third, by disentangling stable between-family differences from dynamic within-family predictive processes, we can establish at which level associations occur, and assess the within-family direction of effects between qualitative involvement and emotional adjustment. Finally, we examined child sex as a moderator of between-family differences in all study variables and within-family predictive effects.

@@

We calculated mean scores for all study variables with multiple indicators. Scale descriptive statistics, number of items, attrition (which ranged from 7-14% across respondents), reliability, and intraclass correlations (ICCs) are reported in Supplementary Table 1; scale correlations are provided in Supplementary Table 2. The difference in fit between structurally invariant models and metrically invariant models was smaller than .01 on all conventional fit indices: RMSEA, CFI, and TLI (see Supplementary Table 3). We thus deemed it defensible to assume metric measurement invariance (i.e., to assume that the factor loadings of each construct are the same across time points and respondents).

@@

To address our research questions, we used multilevel, autoregressive structural equation modeling [@asparouhovDynamicStructuralEquation2018]. In this approach, the variance in qualitative parenting behaviors and emotional adjustment is partialized into stable between-individual differences (the mean level for that individual across all measurement occasions), and within-individual fluctuations (deviations from this person-specific mean). The ICCs (supplementary Table S1) indicate the proportion of variance at the between- versus within-family level. For warmth, consistency, and anger, the ICCs were high, ranging from [.60, .64]. This indicates that about 60% of the variance in these parenting behaviors is due to between-family differences. For inductive reasoning, they were notably lower, [.46, .50]. For emotional adjustment, they were even lower, $r = .25$.

Disentangling between- and within-family level variance allows us to determine at which of these levels associations occur, and to estimate the direction of within-family effects more accurately [@hamakerCritiqueCrosslaggedPanel2015]. Any stable factors which might bias cross-lagged effects - such as the influence of SES or neighborhood effects - are accounted for at the between level, and reflected in the correlations between random intercepts. Secondly, this approach enables us to directly test our hypothesis that between-family differences in quantitative involvement moderate the interplay between qualitative involvement and emotional adjustment: Between-family differences in the interplay between qualitative parenting behaviors and emotional adjustment are estimated as random effects, and we can include between-family level moderators as predictors of these random effects. In this way, we included joint and relative quantitative involvement, and child sex as moderators. Figure \@ref(fig:figure1) shows a graphical representation of the resulting model. We conducted separate analyses for the four qualitative parenting dimensions, because a model including all four variables at once was too complex to estimate.

@@

We hypothesized that warmth, consistency, and inductive reasoning would correlate positively, and anger negatively, with emotional adjustment. At the between-family level, we examined correlations between families' average levels of qualitative parenting behaviors and emotional adjustment (see Table \@ref(tab:tabcors)). These correlations indicate whether between-family differences in the average levels of these parenting behaviors are associated with between-family differences in children's emotional adjustment. Counter to hypotheses $H_{1,2a}$, we found no significant correlations between parental warmth and children's emotional adjustment. In support of hypotheses $H_{1,2b}$, we found significant positive correlations between maternal and paternal consistency, and emotional adjustment. Contrary to hypotheses $H_{1,2c}$, we found that maternal and paternal inductive reasoning were negatively correlated with emotional adjustment. Finally, as hypothesized in $H_{1,2d}$, we found significant negative correlations between maternal and paternal anger, and emotional adjustment. To sum up, in families where parents reported more consistency, children displayed better emotional adjustment - and in families where parents reported more inductive reasoning or anger, children showed poorer emotional adjustment.

@@

We also examined within-family correlations, which are based on within-family fluctuations in the study variables (Table \@ref(tab:tabcors)). These correlations indicate whether fluctuations in parents' parenting behaviors are correlated with fluctuations in their children's emotional adjustment. In support of hypothesis $H_{1a}$, we found significant positive within-family level associations between fathers' and mothers' warmth, and emotional adjustment. This means that if, in any particular wave, parents report more warmth than usual, their children displayed better emotional adjustment than usual. Furthermore, in partial support of hypotheses $H_{1b,c}$, we found significant positive correlations of emotional adjustment with  mothers', but not fathers', consistency, as well as with fathers', but not mothers', inductive reasoning. Finally, contrary to hypothesis $H_{1d}$, we found no significant within-family associations between children's emotional adjustment and maternal or paternal anger. Counter to hypothesis $H_{3}$, none of the differences between mothers and fathers in the strength of either between- or within-family level correlations were significant (see Supplementary Table S4).

## Do within-family fluctuations in parenting predict emotional adjustment or vice versa?

@@

With regard to child effects, in line with hypothesis $H_{2a}$, the average cross-lagged effect of emotional adjustment on maternal warmth was positive and significant (Table \@ref(tab:tabcrosspath)). Thus, on average, when children displayed better emotional adjustment than usual in one wave, their mothers displayed more warmth than usual in the next wave. Contrary to hypothesis $H_{2b}$, the average effect of emotional adjustment on maternal consistency was non-significant. As hypothesized in $H_{2c}$, the intercept for maternal inductive reasoning was positive. Counter to hypothesis $H_{2d}$, the intercept for maternal anger was positive, not negative. Thus, when children displayed better emotional adjustment than usual in one wave, their mothers displayed more anger than usual in the next wave. In support of hypothesis $H_{2a}$, the average child effect of adjustment on paternal warmth was positive and significant. None of the other intercepts ($H_{2b-d}$) for child effects on fathers' parenting were significant, suggesting that fathers' parenting was relatively unresponsive to fluctuations in children's emotional adjustment. The variances of all child effects were significant but small, indicating that the effect of children's emotional adjustment on their parents' child-rearing behaviors differs somewhat between families.

@@

## Does quantitative involvement moderate within-family predictive effects?

We conducted exploratory analyses of interactions between involvement quality and quantity, by investigating whether between-family differences in quantitative involvement had significant effects on the within-family cross-lagged effects between qualitative parenting behaviors and emotional adjustment (Table \@ref(tab:tabinvolvement)). The novelty of this research area precluded us from formulating specific hypotheses. With regard to parenting effects, the only significant moderation we found was for maternal warmth: The positive effect of maternal warmth on adjustment was less pronounced in families in which both parents were highly involved. Specifically, a $+1SD$ increase in joint involvement predicted a $-.14SD$ decrease in the within-families cross-lagged effect of maternal warmth on emotional adjustment (see Table \@ref(tab:tabcrosspath)). This same within-families effect increased by $.11$ of a standard deviation of the effect when relative involvement was one standard deviation higher, i.e., when mothers were more involved than fathers. We found no moderation for paternal warmth, nor for both parents' consistency, inductive reasoning, and anger.

@@
